Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-19 opencl/papers/intro.tex
Nearly at page limit.

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low-level execution and
programming models while allowing the application programmer to adopt
a single high-level programming model.  We show that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.  For algorithms
that can be easily expressed in terms of whole array, data parallel
operations, writing code in Fortran and transforming it automatically
to specific low-level implementations removes the burden from the
programmer of working with tedious, error prone, low-level tools.

In the ideal situation, application programmers would like to adopt a
programming model in which they write their application once and use
automated tools to retarget it to many architectures.  This has proven
to be very challenging historically due to the subtle balance between
high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem
being solved --- in this instance, the abstraction that we care about
are mathematical formulae.  It should not be corrupted with details of
specific target architectures solely for the purpose of single-system
performance.  For certain classes of applications, specifically those
that map well onto a data-parallel programming model, we show that
Fortran 90 contains language features that encourage high-level
programming abstractions without sacrificing performance during
low-level code generation.

@@

However, the performance that these accelerator architectures offer
comes at a cost as processor architectures are trending toward
multiple cores with instances of integrated accelerator units with
user managed memory and less of a reliance on superscalar instruction
level parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  The fundamental question we address is what programming model
and language constructs are best suited to span this set of new
hardware designs.

@@

A common theme amongst new processors is the emphasis on data parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  A proposed approach for programming that
is well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

The problem with many of these choices is that they expose too much
detail about the machine architecture to the programmer.  This is
particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
their codes to fit the threading model used by NVidia GPUs, while
OpenCL requires programmers to provide specially tuned versions of
their code for different classes of machine.  In both cases, the
programmer is also responsible for explicitly managing memory,
including staging of data back and forth from the host CPU and the
accelerator device memory.  While these models have been attractive as
a method for early adopters to utilize these new architectures, they
are less attractive to programmers who do not have the time or
resources to manually port their code to every new architecture and
programming model that emerges.

@@

Transformations for general Fortran procedures are not provided.
Furthermore, a mechanism to transform the calling site to
automatically invoke OpenCL kernels is not provided at this time.
While it is possible to accomplish this task within ROSE, it is
considered outside the scope of this paper.

@@

An initial study was made for an important procedure in PAGOSA, a
non-research, production-grade code at LANL completely written in
data-parallel Fortran.  We investigated automatically transforming
this code to run on LANL's Petaflop Roadrunner computer (a hybrid
mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
source-to-source compiler can automatically vectorize and parallelize
(both within the Cell processor and across nodes) a small section of
this code.  Preliminary results showed a 12 times performance gain on
the Cell processor over a traditional single core processor.  A factor
of 9 is gained from vectorizing and parallelizing the code to run on
the Cell and the rest is gained from automatically overlapping
inter-node communication MPI with computation on the Cell.

%%We study automatic transformations of Fortran for two applications examples.  The first is a simple shallow water model in two dimensions using finite volume methods with .... for time update.  The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.

%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions

@@

Fortran is the oldest high-level programming language in continuous
use since its introduction, and was developed to facilitate the
translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level
program representation to assembly language. Due to its age, it
carries certain arcane baggage that later programming languages have
evolved away from. As a result, Fortran has fallen into disfavor in
certain programming circles.  Modern versions of the language
standardized in 1990, 1995, 2003, and 2008 have removed much of this
legacy baggage, but these changes are not widely known.  Modern
Fortran exhibits features that are similar to other modern programming
languages, and does not mandate the use of legacy features from
decades old, deprecated versions of the language.  Compilers for
Fortran (like those for other languages that have removed features
over time) can prohibit the use of archaic features such as fixed
format code or constructs that have been removed from the language.

With the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. Much of
the arcane baggage associated with it is actually no longer part of
the language used by modern programmers, and exists in a deprecated
form simply to support legacy codes. It is now modular and has many
object-oriented features.  Most importantly for this work, it now
includes a type system in which rich first-class array data types and
corresponding syntax are part of the language, something that
languages like C continue to lack.  Furthermore, Fortran should be of
interest to those studying parallel programming because of its
functional and data parallel constructs and because of the coarray
notation introduced in Fortran 2008. Unlike languages like C and C++,
which are considered to be more modern, Fortran has become a truly
parallel language with features added to recent language standards.

Yet, a likely question that one may pose is ``\emph{Why Fortran and
  not a more modern language like X?}''  The recent rise in interest
in concurrency and parallelism at the language level driven by
multicore CPUs and manycore accelerators has driven a number of new
language developments, both as novel languages and extensions on
existing ones.  For scientific users, new languages and language
extensions to use novel new architectures present a challenge: how do
developers effectively use them while avoiding rewriting their code
and potentially growing dependent on a transient technology that will
vanish tomorrow?

@@

A number of previous efforts have exploited data parallel programming
at the language level to utilize novel architectuers, particularly in
previous decades during the reign of vector and massively parallel
computers in the high performance computing world.  The origin of the
array syntax that was adopted in Fortran 90 can be found in the APL
language\cite{iverson79apl}.  Fortran 90 differed from previous
extensions of Fortran in that parallelism within whole-array
operations was expressed at the expression level instead of via
parallelism within explicit DO-loops (such as within IVTRAN for the
Illiac IV).

>>>>

This paper presents a compiler-level approach for targeting a single
program to multiple, fundamentally different low-level execution
models.  This technique allows the application programmer to adopt
a single high-level programming model without sacrificing performance.
We show that features of
Fortran 90 for data-parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
hardware architectures using low-level programming models such as
OpenCL and CUDA.  For algorithms that can be easily expressed in terms
of whole array, data-parallel operations, writing code in Fortran and
transforming it automatically to specific low-level implementations
removes the burden from the programmer of working with tedious, error
prone, low-level tools.

In the ideal situation, application programmers would like to adopt a
programming model in which they write their application once and use
automated tools to retarget it to many architectures.  This has proven
to be very challenging historically due to the subtle balance between
high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem
being solved --- in this instance, the abstraction that we care about
are mathematical formulae.  It should not be corrupted with details of
specific target architectures solely for the purpose of single-system
performance.  For certain classes of applications, specifically those
that map onto a data-parallel programming model, we show that
Fortran 90 contains language features that encourage high-level
programming abstractions without sacrificing performance during
low-level code generation.

@@

However, the performance that these accelerator architectures offer
comes at a cost, as processor architectures are trending toward
multiple cores with instances of integrated accelerator units (with
user managed memory) and less of a reliance on superscalar instruction
level parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  The fundamental question we address is what programming model
and language constructs are best suited to span this set of new
hardware designs.

@@

A common theme amongst new processors is the emphasis on data parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  The recent CUDA and OpenCL programming
languages are are intended to support this programming model, as are
directive-based methods such as OpenMP or the Accelerator programming
model from the Portland Group~\cite{pgi10accelerator}.

%% Deleted by CER
%%A proposed approach for programming that is well suited to legacy applications and languages are language extensions or libraries that allow programmers to avoid adopting entirely new languages.

The problem with many of these choices is that they expose too much
detail about the machine architecture to the programmer.  This is
particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
their codes to fit the threading model used by NVidia GPUs, while
OpenCL requires programmers to provide specially tuned versions of
their code for different classes of machine.  In both cases, the
programmer is responsible for explicitly managing memory,
including staging of data back and forth from the host CPU and the
accelerator device memory.  While these models have been attractive as
a method for early adopters to utilize these new architectures, they
are less attractive to programmers who do not have the time or
resources to manually port their code to every new architecture and
programming model that emerges.

@@

Transformations for general Fortran procedures are not attempted.
Furthermore, a mechanism to transform the calling site to
automatically invoke OpenCL kernels is not provided at this time.
While it is possible to accomplish this task within ROSE, it is
considered outside the scope of this paper.

@@

An initial study was made for an important procedure in PAGOSA, a
non-research, production-grade code at LANL completely written in
data-parallel Fortran.  We investigated automatically transforming
this code to run on LANL's Petaflop Roadrunner computer (a hybrid
mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
source-to-source compiler can automatically vectorize and parallelize
a small section of this code for the Cell processor.  Preliminary results
showed a 9 times performance gain of the transformed code when compared
with the original serial version on a traditional single-core processor.

@@

Fortran is the oldest high-level programming language in continuous
use since its introduction, and was developed to facilitate the
translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level
program representation to assembly language. Due to its age, it
carries certain arcane baggage.  However
with the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. 
It is now modular and has many
object-oriented features.  Most importantly for this work, it now
includes a type system in which rich first-class array data types and
corresponding syntax are part of the language, something that
languages like C continue to lack.  Furthermore, Fortran should be of
interest to those studying parallel programming because of its
functional and data parallel constructs and because of the coarray
notation introduced in Fortran 2008. Unlike languages like C and C++,
Fortran has become a truly
parallel language with features added to recent language standards.

%% - deleted by CER
%% that later programming languages have evolved away from. As a result, Fortran has fallen into disfavor in certain programming circles.  Modern versions of the language standardized in 1990, 1995, 2003, and 2008 have removed much of this legacy baggage, but these changes are not widely known.  Modern Fortran exhibits features that are similar to other modern programming languages, and does not mandate the use of legacy features from decades old, deprecated versions of the language.  Compilers for Fortran (like those for other languages that have removed features over time) can prohibit the use of archaic features such as fixed format code or constructs that have been removed from the language.

Yet, a likely question that one may pose is ``\emph{Why Fortran and
  not a more modern language like X?}''  The recent rise in interest
in concurrency and parallelism at the language level driven by
multicore CPUs and manycore accelerators has driven a number of new
language developments, both as novel languages and extensions on
existing ones.  For scientific users, new languages and language
extensions to use novel new architectures present a challenge: how do
developers effectively use them while avoiding rewriting code
and potentially growing dependent on a transient technology that will
vanish tomorrow?

@@

A number of previous efforts have exploited data parallel programming
at the language level to utilize novel architectures, particularly in
previous decades during the reign of vector and massively parallel
computers in the high performance computing world.  The origin of the
array syntax that was adopted in Fortran 90 can be found in the APL
language\cite{iverson79apl}.  Fortran 90 differed from previous
extensions of Fortran in that parallelism within whole-array
operations was expressed at the expression level instead of via
parallelism within explicit DO-loops (such as within IVTRAN for the
Illiac IV).

### 2010-12-19 opencl/papers/programming-model.tex
Nearly at page limit.

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a subset of Fortran that
employs a data-parallel programming model.  In particular, it
encourages use of four language features introduced in Fortran 90 ---
array notation, shift functions, elemental functions, and pure
procedures.  From these language constructs, we are able to easily
transform them to a lower-level CUDA or OpenCL implementation.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements that are in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.  This has the
benefit of avoiding explicit looping in the code and maintaining a
high-level style that is closer to the original mathematical
specification of the problem being solved.  More importantly from a
compilation perspective, this defers decisions about how to implement
these whole-array operations to a compilation tool.  Sophisticated
loop analysis to identify parallelism within the sequential loop code
are not necessary.  When faced with a novel architecture such as
modern GPUs that are ideally suited to data parallel programming
models, the fit between Fortran arrays and these systems is quite
clean.

A Fortran array combines a base pointer to memory with metadata that
describes the array shape, stride and size. It should be noted that
other languages that provide rich array data types and whole array
operations may also be suitable targets for the transformations
described in this paper.  Array variables can be used in expressions
based on whole-array operations.  For example, if {\tt A}, {\tt B},
and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
scalar, then the statement

@@

Many array-based algorithms require the same operation to be performed
on each element of the array using the value of that element and some small
set of neighboring cells.  Often programmers implement these operations that
are local to each element within the inner-loop of a set of nested FOR-
or DO-loops using offsets relative to the current array index.  An alternative
to this local-view of the algorithm is to take a global view and write the
algorithm in terms of the whole array.  For example, consider a 1D array in
which we wish to subtract the $(i-1)$th element from the $i$th for all
elements.  One way to look at this is that we are subtracting the entire array
shifted by one element from itself.

Fortran provides a set of shifting operators that allow programmers to
define operations based on shifted arrays.  These intrinsic operators take
an array, a dimension, and the amount by which it should be shifted (using
the sign to indicate direction).  By defining operations on entire arrays
based on a global view of them shifted relative to each other, programmers can
avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
analysis of the extent of the set of shifted arrays in a given expression
allows analysis tools to determine the amount of temporary or buffer storage
necessary to hold intermediate values during whole array operations.  With
explicit loops, programmers must maintain this temporary storage manually.

@@

Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
Fortran, regions are expressed as pointers to a subsection of an existing array.
Regions may refer to an interior portion of an array or the entire array.

The use of regions implies the use of the halo (or ghost) cell pattern 
where the size of an array is increased to provide extra array elements
surrounding the interior portion of the array.  
%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf

Regions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
nor the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

@@

Unlike OpenCL, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

@@

\subsection{Parallelism}

% TODO - Matt - do you want a try at this.  Not sure we have room for
% this subsection. I briefly mention node parallelism in other sections.

% need to describe processor hierarchy of nodes and accelerators
% need to describe memory hierarchy as well in regards to global array view,
% node array view (with a copy is GPU global memory) and local GPU memory which
% further subdivides the node array memory

% figures on slicing and dicing of memory region go here

@@

%% NEW - CER
An example of this style of programming in Fortran is shown in

{\small
\begin{verbatim}
  Bz = Bz &
     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
\end{verbatim}
}

%% NEW - CER

This example is a solution to Maxwell's equations for the $z$
component of the magnetic field using Fortran array syntax.  Note that
there are no explicit loops in this example.  The operators {\tt +},
{\tt -}, and {\tt *} are applied to all of the elements of the
three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey},
individually.  This is why data parallelism has been called
collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhye93}.  As the {\tt cshift}
function and the array-valued expressions all semantically returns a
value, this style of programming is also similar to functional
programming (or value-oriented
programming).

@@

%% NEW - CER
This style of programming meets the requirements we have set for a programming
model for developing applications suitable for acceleration.  It allows the programmer to
program at a very-high level of abstraction while providing the compiler with
maximum flexibility in targeting the application for a particular hardware
architecture.  The data parallel programming model simultaneously meets the
seemingly conflicting goals of maintainability, portability, and performance.

%% NEW - CER
Unfortunately, this style of programming has never really caught on because
when Fortran 90 was first introduced, performance was relatively poor and thus
programmers shied away from using array syntax (even now, some are actively
counseling against its usage because of performance issues~\cite{Levesque:SC08}).
Thus the Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.
A goal of this paper is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on
accelerator architectures.

\subsection{Limitations}

Only Fortran procedures are transformed into OpenCL kernels.  The programmer
must currently explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible using ROSE to modify the calling
site so that the entire program can be transformed but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.

@@

Array sizes must be multiples of the local kernel size, {\tt
  get\_local\_size(0)*get\_local\_size(1)}.  This may be relaxed in
the future.

>>>>

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a language subset that
employs a data-parallel programming model.  In particular, it
encourages use of four language features introduced in Fortran 90 ---
array notation, shift functions, elemental functions, and pure
procedures.  From these language constructs, we are able to easily
transform them to a lower-level CUDA or OpenCL implementation.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.
Array variables can be used in expressions
based on whole-array operations.  For example, if {\tt A}, {\tt B},
and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
scalar, then the statement

%% Deleted by CER
%%This has the benefit of avoiding explicit looping in the code and maintaining a high-level style that is closer to the original mathematical specification of the problem being solved.  More importantly from a compilation perspective, this defers decisions about how to implement these whole-array operations to a compilation tool.  Sophisticated loop analysis to identify parallelism within the sequential loop code are not necessary.  When faced with a novel architecture such as modern GPUs that are ideally suited to data parallel programming models, the fit between Fortran arrays and these systems is quite clean.

%% Deleted by CER
%% A Fortran array combines a base pointer to memory with metadata that describes the array shape, stride and size. It should be noted that other languages that provide rich array data types and whole array operations may also be suitable targets for the transformations described in this paper.

@@

Many array-based algorithms require the same operation to be performed
on each element of the array using the value of that element and some small
set of neighboring cells. As described above, this can often be done with
shift operations.
Fortran provides a set of shifting operators that allow programmers to
define operations based on shifted arrays.  These intrinsic operators take
an array, a dimension, and the amount by which it should be shifted (using
the sign to indicate direction).  By defining operations on entire arrays
based on a global view of them shifted relative to each other, programmers can
avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
analysis of the extent of the set of shifted arrays in a given expression
allows analysis tools to determine the amount of temporary or buffer storage
necessary to hold intermediate values during whole array operations.  With
explicit loops, programmers must maintain this temporary storage manually.

%% Deleted by CER
%% Often programmers implement these operations that are local to each element within the inner-loop of a set of nested FOR- or DO-loops using offsets relative to the current array index.  An alternative to this local-view of the algorithm is to take a global view and write the algorithm in terms of the whole array.  For example, consider a 1D array in which we wish to subtract the $(i-1)$th element from the $i$th for all elements.  One way to look at this is that we are subtracting the entire array shifted by one element from itself.

@@

Borrowing ideas from ZPL, we introduce region functions to Fortran.
In Fortran, these functions return a copy of an
existing array or array section.  This is unlike ZPL where regions
are first-class citizens and have special syntax.
These functions imploy a halo (or ghost) cell pattern,
where the size of an array is increased to provide extra array elements
surrounding the interior portion of the array.
%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf

Region functions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
nor the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

@@

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

@@

\subsection{Advantages}

@@

%% Deleted by CER
%% An example of this style of programming in Fortran is shown in
%
%{\small
%\begin{verbatim}
%  Bz = Bz &
%     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
%     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
%\end{verbatim}
%}

%%This example is a solution to Maxwell's equations for the $z$ component of the magnetic field using Fortran array syntax.  Note that there are no explicit loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey}, individually.

Data parallelism has been called
collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhye93}.  As the {\tt cshift}
function and the array-valued expressions all semantically returns a
value, this style of programming is also similar to functional
programming (or value-oriented programming).

@@

%% Deleted by CER
%% This style of programming meets the requirements we have set for a programming model for developing applications suitable for acceleration.  It allows the programmer to program at a very-high level of abstraction while providing the compiler with maximum flexibility in targeting the application for a particular hardware architecture.  The data parallel programming model simultaneously meets the seemingly conflicting goals of maintainability, portability, and performance.

Unfortunately, this style of programming has never really caught on because
when Fortran 90 was first introduced, performance was relatively poor and thus
programmers shied away from using array syntax (even now, some are actively
counseling against its usage because of performance issues~\cite{Levesque:SC08}).
Thus the Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.
A goal of this paper is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on
accelerator architectures.

\subsection{Current Limitations}

Only Fortran procedures are transformed into OpenCL kernels.  The programmer
must explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible, using ROSE, to modify the calling
site so that the entire program can be transformed, but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively
new to ROSE, some generated code must be modified.  For example, the
{\tt \_\_global} attribute for kernel arguments was added by hand.

It is assumed that memory for all arrays reside on the device.  The programmer must
copy memory to and from the device.  In addition, array size (neglecting ghost cell
regions) must be multiples of the global OpenCL kernel size.

@@

### 2010-12-19 opencl/papers/shallow-water.tex
fixed spelling error
Nearly at page limit.

where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than $dt\leqq(\sqrt{gh}+|u|)/dx$.

@@

The Fortran kernel procedure {\tt wave\_advance} for the shallow water code
is declared as:

@@

where {\tt H, U}, and {\tt V} are start variables for height and
$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.

Temporary arrays are required for the interior copies {\tt iH, iU}, and
{\tt iV} of the state variables and for the flux quantities {\tt Hx, Hy, Ux,
Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,

@@

Halos variables for the interior and cell faces are defined as

@@

Note that the halos for the four faces each have a 0 in the
definition. Thus the returned array copy will have size that is larger
than the interior regions that use {\tt [1,1,1,1]}.  This is because
there is one more cell face quantity than there are cells in a given
direction.

The Lax-Wendroff update algorithm for the shallow-water model has
two steps.  The fluxes are first updated on cell faces, for
example,

@@

\noindent
updates the $x$-direction flux for the height state variable.

Then second step uses these fluxes to update the interior regions for
the state variables.  For example,

@@

\noindent
Note that face halos have been redefined so that the array copy
returned by now has the same size as the interior region.

These simple code segments show how the shallow water model is
implemented in standard Fortran (2003) using the data-parallel
programming model described above.  The resulting code is simple,
consise, and easy to understand.  However it does \emph{not}
necessarily perform well because of the temporary array variables,
especially those produced by the {\tt region} function.  This is
generally true of algorithms that use Fortran shift functions as well,
as some Fortran compilers (e.g., gfortran) do not generate optimal
code for shifts.  We note (as seen below) that these temporary array
copies are replaced by scalars in the transformed Fortran code so
there is no performance penalty for using data-parallel statements as
outlined.

>>>>

\noindent
where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$.

@@

The coding style to implement the serial version of the shallow water model
is now shown.  The Fortran kernel procedure {\tt wave\_advance} 
is declared as:

@@

\noindent
where {\tt H, U}, and {\tt V} are state variables for height and
$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.

Temporary arrays are required for the interior copies {\tt iH, iU}, and
{\tt iV} of the state variables and for the quantities {\tt Hx, Hy, Ux,
Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,

@@

Halo variables for the interior and cell faces are

@@

\noindent
Note that the halos for the four faces each have a 0 in the
definition. Thus the returned array copy will have size that is larger
than the interior regions that use {\tt [1,1,1,1]}.  This is because
there is one more cell face quantity than there are cells in a given
direction.

The first Lax-Wendroff step updates state variables on the cell
faces, for example,

@@

\noindent
updates the $x$-direction height variable.  The second step then uses
these face variables to update the interior region, for example,

@@

\noindent
Note that face halos have been redefined so that the array copy
returned has the same size as the interior region.

These simple code segments show how the shallow water model is
implemented in standard Fortran (2003) using the data-parallel
programming model described above.  The resulting code is simple,
concise, and easy to understand.  However it does \emph{not}
necessarily perform well because of the temporary array variables,
especially those produced by the {\tt region} function.  This is
generally true of algorithms that use Fortran shift functions as well,
as some Fortran compilers (e.g., gfortran) do not generate optimal
code for shifts.  We note (as seen below) that these temporary array
copies are replaced by scalars in the transformed Fortran code so
there is no performance penalty for using data-parallel statements as
outlined.

### 2010-12-19 opencl/papers/source-to-source.tex
Nearly at page limit.

Device memory is separated hierarchically.  A thread instance has access to
its own memory, thread groups to OpenCL local memory, and all thread groups
have access to OpenCL global memory.  When multiple members of a thread group
access the same memory elements, for example if {\tt region} or {\tt shift}
functions are called, for performance reasons it is often best if global
memory accessed by a thread group is copied into local memory.

@@

This sections outlines the OpenCL equivalent syntax for portions of the
Fortran shallow-water code described in the previous section.  The notation
uses uppercase for arrays and lowercase for scalar quantities.  Interior and
region array copies are denoted by an i or an r preceeding the array.  For
example, {\tt iH = interior(H, halo)} is a local copy of interior region of
array {\tt H} (representing height) in the shallow water code.

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt \_\_local float
  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are copied
to local memory by the compiler with inlined library functions.  The flux
array temporaries (region variables) are declared similarly but initialized to
zero with inlined library functions.  Interior variables are simple scalars,
e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
regions are shifted and thus \emph{shared} by threads within a thread group.

@@

Region variables are more complicated because they are arrays.

@@

where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL compiler.

>>>>

Device memory is separated hierarchically.  A thread instance has access to
its own memory, thread groups to OpenCL local memory, and all thread groups
have access to OpenCL global memory.  When multiple members of a thread group
access the same memory elements (if {\tt region} or {\tt shift}
functions are called), for performance reasons it is often best if global
memory accessed by a thread group is copied into local memory.

@@

This sections outlines the OpenCL equivalent syntax for portions of the
Fortran shallow-water code described previously.  The notation
uses uppercase for arrays and lowercase for scalar quantities.  Interior
variables are denoted by an i preceeding the array name.  For
example, {\tt iH = interior(H, halo)} is a local copy of interior region of
array {\tt H} (representing height) in the shallow water model.

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt \_\_local float
  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are
copied to local memory by the compiler with inlined library functions.  The
array temporaries defined on cell faces are declared similarly while interior
variables are simple scalars, e.g., {\tt float iH;}.  Region variables cannot
be scalar objects because regions may be shifted and thus \emph{shared} by
threads within a thread group.

@@

\noindent
Region variables are more complicated because they are arrays.

@@

\noindent
where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL compiler.
